---
title: "Exploring PCA Structure"
author: "David Hughes"
output: html_document
---

```{r}
source("parameter_files/10.parameters.R")
library(tidyverse)
```


## Load the data

```{r}
load(rdata_file)
clindata = read.table(clin_file, header = TRUE, sep = ",", as.is = TRUE)
```

## Merge the data

```{r}
m = match(qc_data$sample_data[, "CLIENT_SAMPLE_ID"], clindata[,"Client.Sample.ID."])
clindata_m = clindata[m,]
```


## Plot the PCA

### Is there obvious structure by Source Study ?

```{r, fig.width = 7, fig.height = 6}
qc_data$sample_data %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(color = SOURCE_STUDY) ) +
  theme_bw()
```

#### How many of each source study are there?

```{r}
table(qc_data$sample_data$SOURCE_STUDY)
```

### run k-means on PC1 and PC2

```{r}
k = kmeans(qc_data$sample_data[, c("PC1","PC2")], centers = 2)
qc_data$sample_data$k = as.factor( k$cluster )
table(qc_data$sample_data$k)
```

#### Run a PCA with the k-means structure

```{r, fig.width = 7, fig.height = 6}
qc_data$sample_data %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(color = as.factor(k) ) ) +
  theme_bw()
```

There are ~141 outliers.


#### Is there obvious structure by run platform?

```{r, fig.width = 7, fig.height = 6}
qc_data$sample_data %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(color = as.factor(POLAR) ) ) +
  theme_bw()
```

Run structure of box number and platform

```{r}
table( qc_data$sample_data$BOX_NUMBER, qc_data$sample_data$NEG )
# table( qc_data$sample_data$BOX_NUMBER, qc_data$sample_data$polar )
# table( qc_data$sample_data$BOX_NUMBER, qc_data$sample_data$posearly )
# table( qc_data$sample_data$BOX_NUMBER, qc_data$sample_data$poslate )
```

## how many samples from each clinic ?

```{r}
# table( clindata$timepoint )
# table( clindata$source_study )
table( clindata$site )
```


#### Is there obvious structure by site ?

```{r, fig.width = 7, fig.height = 6}

qc_data$sample_data$site = as.factor(clindata_m$site)

levels(qc_data$sample_data$site) <- c("G","C","I","J","K","D","L","H","A","B","F","ALSPAC")

w = which(qc_data$sample_data$SOURCE_STUDY == "alspac")
qc_data$sample_data$site[w] = "ALSPAC"
qc_data$sample_data$site <- ordered(qc_data$sample_data$site, levels = c("A","B","C","D","F","G","H","I","J","K","L","ALSPAC"))
###
xl = paste0("PC1 (", round(qc_data$varexp[1,1] * 100,digits=1), "%)"  )
yl = paste0("PC2 (", round(qc_data$varexp[2,1] * 100,digits=1), "%)"  )

qc_data$sample_data %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(fill = site ), shape = 21, size = 4, alpha = 0.7 ) +
  theme_bw() +
  labs(fill = "Site", x = xl, y = yl)
ggsave(paste0(fig_dir,"Fig2_PC_structure.jpeg"),dpi=600)
```


```{r}
table(clindata_m$site, clindata_m$timepoint)
```

## Grab IDs for the siteG samples that are outliers

```{r}
outlier_data = qc_data$sample_data %>% filter(site == "G" & PC1 > 0  & PC2 < -5)
#f = paste0(results_dir, "siteG_sample_outlier_anno.txt")
#write.table(outlier_data, file = f, col.names = TRUE, row.names = TRUE, sep = "\t", quote = FALSE)
```


```{r}
# site_timepoint = paste0(clindata_m$site, "_", clindata_m$timepoint)
site_timepoint = sapply(1:nrow(clindata_m), function(i){
  paste0(clindata_m$site[i], "_", clindata_m$timepoint[i])
})
w = grep("BIR", site_timepoint)
site_timepoint[-w] = "other"
qc_data$sample_data$site_timepoint = site_timepoint
```


```{r}
qc_data$sample_data %>% arrange(desc(site_timepoint)) %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(fill = as.factor(site_timepoint) ), shape = 21, size = 4, alpha = 0.7 ) +
  theme_bw() +
  labs(fill = "site_timepoint", x = xl, y = yl)
```


```{r}
table(clindata_m$site, clindata_m$sex)
```



```{r}
# site_timepoint = paste0(clindata_m$site, "_", clindata_m$timepoint)
site_sex = sapply(1:nrow(clindata_m), function(i){
  paste0(clindata_m$site[i], "_", clindata_m$sex[i])
})
w = grep("BIR", site_timepoint)
site_sex[-w] = "other"
qc_data$sample_data$site_sex = site_sex
```


```{r}
qc_data$sample_data %>% arrange(desc(site_sex)) %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(fill = as.factor(site_sex) ), shape = 21, size = 4, alpha = 0.7 ) +
  theme_bw() +
  labs(fill = "site_sex", x = xl, y = yl)
```


```{r, fig.height = 9, fig.width = 8}
# names(qc_data)
library(patchwork)
##
p1 =qc_data$sample_data %>% ggplot(aes(x = site, y = sample_missingness)) +
  geom_boxplot(aes(fill = site)) + theme_bw()

p2 = qc_data$sample_data %>% ggplot(aes(x = site, y = sample_missingness_w_exclusions)) +
  geom_boxplot(aes(fill = site)) + theme_bw()

p3 = qc_data$sample_data %>% ggplot(aes(x = site, y = TPA_total)) +
  geom_boxplot(aes(fill = site)) + theme_bw()

p4 = qc_data$sample_data %>% ggplot(aes(x = site, y = TPA_completefeature)) +
  geom_boxplot(aes(fill = site)) + theme_bw()

( p1 / p2 / p3 / p4 ) + plot_layout(guides = "collect")
```



```{r}
year = sapply(clindata_m$visitdate, function(x){strsplit(x, split = "-")[[1]][1]})

# site_timepoint = paste0(clindata_m$site, "_", clindata_m$timepoint)
site_year = sapply(1:nrow(clindata_m), function(i){
  paste0(clindata_m$site[i], "_", year[i])
})
w = grep("BIR", site_year)
site_year[-w] = "other"
qc_data$sample_data$site_year = site_year
```


```{r, fig.width = 7, fig.height = 6}
qc_data$sample_data %>% arrange(desc(site_year)) %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point( aes(fill = as.factor(site_year) ), shape = 21, size = 4, alpha = 0.7 ) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  labs(fill = "site_year", x = xl, y = yl)
```


