---
title: "Comparing results from old and new siteB data"
author: "Maddy"
initiation_date: "13/04/2023"
output:
  html_document: default
---

<style type="text/css">
div.main-container {
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r global_options, echo = FALSE, include = FALSE}
options(width = 999)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
```

```{r setup, results='asis', echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# read in parameter file
source("parameter_files/17.parameters.R")

# load libraries
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(gtools)
library(qqman)
library(ggpubr)
library(ggrepel)
```


```{r read in results, echo=FALSE}
old <- readRDS(file=paste0(siteB_dir,"05_mixed_results.rds"))
old <- old$rnt_imputed_adj_as$metabolon
new <- readRDS(file=paste0(results_output_dir, "05_mixed_results.rds"))
new <- new$rnt_imputed_adj_as
  
results <- list(old=old, new=new)

# read in feature data and median fold change results
feature_data <- read.csv(paste0(data_input_dir,"bbs_alspac_outliers_removed_2025_04_04_Metabolon_featuredata.txt"), sep = "\t", header=T)

# remove unwanted columns - only keep n, beta, se, p
for (i in 1:length(results)) {
    n_col <- colnames(results[[i]])[(which(grepl("n_", colnames(results[[i]]))==T))]
    beta_col <- colnames(results[[i]])[(which(grepl("beta_", colnames(results[[i]]))==T))]
    se_col <- colnames(results[[i]])[(which(grepl("se_", colnames(results[[i]]))==T))]
    p_col <- "P_mixed"
    results[[i]] <- results[[i]][,c(n_col,beta_col,se_col,p_col)]
  }


# merge in feature data to each set of results
for (i in 1:length(results)) {
    results[[i]] <- merge(results[[i]], feature_data, by.y = "feature_names", by.x = "row.names", sort = F)
  }
```

### top 10 table, volcano plot & scatters comparing betas and p-vals
#### 1. Old siteB data
#### 2. New siteB data
```{r summary loop, echo=FALSE, eval=TRUE}
options(width = 250)
for (i in 1:length(results)) {
  print(names(results)[i])
    # top 10 hits
    n_col <- colnames(results[[i]])[(which(grepl("n_", colnames(results[[i]]))==T))]
    beta_col <- colnames(results[[i]])[(which(grepl("beta_", colnames(results[[i]]))==T))]
    se_col <- colnames(results[[i]])[(which(grepl("se_", colnames(results[[i]]))==T))]
    p_col <- colnames(results[[i]])[(which(grepl("P_", colnames(results[[i]]))==T))]
    print(head(results[[i]][,c("Row.names", n_col, beta_col, se_col, p_col, "SUPER_PATHWAY", "SUB_PATHWAY", "PLOT_NAME")], 10))
    # volcano
    p_col <- results[[i]][(which(grepl("P_", colnames(results[[i]]))==T))]
    p_col <- unlist(p_col)
    volcano <- ggplot(data=results[[i]], 
                  aes(x=beta_mixed, y=-1*log10(p_col), col=SUPER_PATHWAY)) + 
    geom_point() +
    theme_minimal() + 
    geom_vline(xintercept=0, col="red") +
    geom_hline(yintercept=-log10(0.05/446), col="red") + 
    xlab("beta") +
    ylab(expression("-log"[10]*"(p-value)"))
    print(volcano)
    
    # comparison scatters - compare best (2) with new (1)
    if (i==1) {
      # make temp df of betas
      compare_beta <- as.data.frame(matrix(nrow = nrow(feature_data)))
      rownames(compare_beta) <- feature_data$feature_names
      compare_beta <- merge(compare_beta, results[[1]], by.x = "row.names", by.y = "Row.names")[,c("Row.names","beta_mixed")]
      colnames(compare_beta)[2] <- paste0(names(results[1]), ".rnt_imputed_adj_as")
      compare_beta <- merge(compare_beta, results[[2]], by = "Row.names")[,c("Row.names", "old.rnt_imputed_adj_as", "beta_mixed")]
      colnames(compare_beta)[3] <- paste0(names(results[2]), ".rnt_imputed_adj_as")
      # scatter plot of betas
      lm_b <- summary(lm(data=compare_beta, new.rnt_imputed_adj_as ~ old.rnt_imputed_adj_as))
      p1 <- ggplot(compare_beta, aes( x = old.rnt_imputed_adj_as, y = new.rnt_imputed_adj_as )) +
        geom_point(size = 1.5) +
        geom_smooth(method = 'lm', formula = y ~ x) +
        theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
        geom_label_repel(aes(label = ifelse((new.rnt_imputed_adj_as<(-0.7) & old.rnt_imputed_adj_as>0.2), Row.names, "")), size = 3) +
        xlim(-2, 2) + ylim(-2,2) +
        xlab(names(results[1])) + ylab(names(results[2])) +
        annotate("text", x = -1, y = 1.6, label = (paste0("R^2 = ", lm_b$adj.r.squared)), size = 5) +
        ggtitle("betas")
      print(p1)
      print("compid_36803 is 3beta,7alpha-dihydroxy-5-cholestenoate, a sterol (lipid superpathway), not associated with surgery in either new or old dataset")
      # make temp df of p-vals
      compare_p <- as.data.frame(matrix(nrow = nrow(feature_data)))
      rownames(compare_p) <- feature_data$feature_names
      compare_p <- merge(compare_p, results[[1]], by.x = "row.names", by.y = "Row.names")[,c("Row.names","P_mixed")]
      colnames(compare_p)[2] <- paste0(names(results[1]), ".rnt_imputed_adj_as")
      compare_p <- merge(compare_p, results[[2]], by = "Row.names")[,c("Row.names", "old.rnt_imputed_adj_as", "P_mixed")]
      colnames(compare_p)[3] <- paste0(names(results[2]), ".rnt_imputed_adj_as")
      # scatter plot of ps
      lm_p <- summary(lm(data=compare_p, new.rnt_imputed_adj_as ~ old.rnt_imputed_adj_as))
      p1 <- ggplot(compare_p, aes( x = -1*log10(old.rnt_imputed_adj_as), y = -1*log10(new.rnt_imputed_adj_as) )) +
        geom_point(size = 1.5) +
        geom_smooth(method = 'lm', formula = y ~ x) +
        theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
        xlab(names(results[1])) + ylab(names(results[2])) +
        annotate("text", x = 5, y = 11, label = (paste0("R^2 = ", lm_p$adj.r.squared)), size = 5) +
        ggtitle("-1*log10(p-val)")
      print(p1)
    }
}

```


#### Some of the differences in the top 10s are valine, glycine, 3-methyl-2-oxobutyrate, oleoyl choline, citrulline
#### Let's look at the raw data for these metabs...
```{r compare raw data, echo=FALSE}
# read in the two datasets to compare the raw data for some metabs - can compare raw and rnt'd
# new data
load(paste0(data_output_dir, "intermediate/01_dataset_list.RData"))
## keep only siteB samples
siteB_mainfest <- read.csv(siteB_manifest)
siteB <- siteB_mainfest[,"Study.Identifier"]
for (i in 1:length(dataset_list)) {
  w <- which(dataset_list[[i]]$studyid %in% siteB)
  dataset_list[[i]] <- dataset_list[[i]][w,]
}
new_raw <- dataset_list$raw
new_rnt <- dataset_list$rnt_raw

# old data
old <- readRDS(file=paste0(old_siteB_dir,"01_dataset_list.rds"))
old_raw <- old$raw$metabolon
old_rnt <- old$rnt_imputed$metabolon

# sample IDs might not match (different tubes) so create a studyID/timepoint var to match on (need to make 36 months end instead)
old_raw$Visit.Identifier <- gsub("36 months", "end", old_raw$Visit.Identifier)
old_rnt$Visit.Identifier <- gsub("36 months", "end", old_rnt$Visit.Identifier)
old_raw$id_timepoint <- paste0(old_raw$Study.Identifier, "_", old_raw$Visit.Identifier)
old_rnt$id_timepoint <- paste0(old_rnt$Study.Identifier, "_", old_rnt$Visit.Identifier)

# now create temp dfs to allow comparison of selected metabs
## VALINE - compid_1649
compare_valine_raw <- merge(new_raw[,c("id_timepoint","compid_1649")], old_raw[,c("id_timepoint","compid_1649")], by = "id_timepoint")
compare_valine_rnt <- merge(new_rnt[,c("id_timepoint","compid_1649")], old_rnt[,c("id_timepoint","compid_1649")], by = "id_timepoint")

lm_val <- summary(lm(data=compare_valine_raw, compid_1649.x ~ compid_1649.y))
p1 <- ggplot(compare_valine_raw, aes( x = compid_1649.x, y = compid_1649.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = 1, y = 2, label = (paste0("R^2 = ", lm_val$adj.r.squared)), size = 5) +
    ggtitle("valine_raw")
print(p1)

lm_val <- summary(lm(data=compare_valine_rnt, compid_1649.x ~ compid_1649.y))
p1 <- ggplot(compare_valine_rnt, aes( x = compid_1649.x, y = compid_1649.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = -1, y = 2.5, label = (paste0("R^2 = ", lm_val$adj.r.squared)), size = 5) +
    ggtitle("valine_rnt")
print(p1)


## 3-methyl-2-oxybutyrate - compid_44526
compare_3methox_raw <- merge(new_raw[,c("id_timepoint","compid_44526")], old_raw[,c("id_timepoint","compid_44526")], by = "id_timepoint")
compare_3methox_rnt <- merge(new_rnt[,c("id_timepoint","compid_44526")], old_rnt[,c("id_timepoint","compid_44526")], by = "id_timepoint")

lm_3m <- summary(lm(data=compare_3methox_raw, compid_44526.x ~ compid_44526.y))
p1 <- ggplot(compare_3methox_raw, aes( x = compid_44526.x, y = compid_44526.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = 1, y = 2, label = (paste0("R^2 = ", lm_3m$adj.r.squared)), size = 5) +
    ggtitle("3methox_raw")
print(p1)

lm_3m <- summary(lm(data=compare_3methox_rnt, compid_44526.x ~ compid_44526.y))
p1 <- ggplot(compare_3methox_rnt, aes( x = compid_44526.x, y = compid_44526.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = -0.5, y = 2.5, label = (paste0("R^2 = ", lm_3m$adj.r.squared)), size = 5) +
    ggtitle("3methox_rnt")
print(p1)


## GLYCINE - compid_58
compare_glycine_raw <- merge(new_raw[,c("id_timepoint","compid_58")], old_raw[,c("id_timepoint","compid_58")], by = "id_timepoint")
compare_glycine_rnt <- merge(new_rnt[,c("id_timepoint","compid_58")], old_rnt[,c("id_timepoint","compid_58")], by = "id_timepoint")

lm_gly <- summary(lm(data=compare_glycine_raw, compid_58.x ~ compid_58.y))
p1 <- ggplot(compare_glycine_raw, aes( x = compid_58.x, y = compid_58.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = 1, y = 2, label = (paste0("R^2 = ", lm_gly$adj.r.squared)), size = 5) +
    ggtitle("glycine_raw")
print(p1)

lm_gly <- summary(lm(data=compare_glycine_rnt, compid_58.x ~ compid_58.y))
p1 <- ggplot(compare_glycine_rnt, aes( x = compid_58.x, y = compid_58.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = -1, y = 2.5, label = (paste0("R^2 = ", lm_gly$adj.r.squared)), size = 5) +
    ggtitle("glycine_rnt")
print(p1)


## Oleoyl choline - compid_53260
compare_ochol_raw <- merge(new_raw[,c("id_timepoint","compid_53260")], old_raw[,c("id_timepoint","compid_53260")], by = "id_timepoint")
compare_ochol_rnt <- merge(new_rnt[,c("id_timepoint","compid_53260")], old_rnt[,c("id_timepoint","compid_53260")], by = "id_timepoint")

lm_ochol <- summary(lm(data=compare_ochol_raw, compid_53260.x ~ compid_53260.y))
p1 <- ggplot(compare_ochol_raw, aes( x = compid_53260.x, y = compid_53260.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = 1, y = 2, label = (paste0("R^2 = ", lm_ochol$adj.r.squared)), size = 5) +
    ggtitle("ochol_raw")
print(p1)

lm_ochol <- summary(lm(data=compare_ochol_rnt, compid_53260.x ~ compid_53260.y))
p1 <- ggplot(compare_ochol_rnt, aes( x = compid_53260.x, y = compid_53260.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = -1, y = 2.5, label = (paste0("R^2 = ", lm_ochol$adj.r.squared)), size = 5) +
    ggtitle("ochol_rnt")
print(p1)


## citrulline - compid_2132
compare_citrulline_raw <- merge(new_raw[,c("id_timepoint","compid_2132")], old_raw[,c("id_timepoint","compid_2132")], by = "id_timepoint")
compare_citrulline_rnt <- merge(new_rnt[,c("id_timepoint","compid_2132")], old_rnt[,c("id_timepoint","compid_2132")], by = "id_timepoint")

lm_cit <- summary(lm(data=compare_citrulline_raw, compid_2132.x ~ compid_2132.y))
p1 <- ggplot(compare_citrulline_raw, aes( x = compid_2132.x, y = compid_2132.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = 1, y = 2, label = (paste0("R^2 = ", lm_cit$adj.r.squared)), size = 5) +
    ggtitle("citrulline_raw")
print(p1)

lm_cit <- summary(lm(data=compare_citrulline_rnt, compid_2132.x ~ compid_2132.y))
p1 <- ggplot(compare_citrulline_rnt, aes( x = compid_2132.x, y = compid_2132.y) ) +
    geom_point(size = 1.5) +
    geom_smooth(method = 'lm', formula = y ~ x) +
    theme(axis.text=element_text(size=10), axis.title=element_text(size=12,face="bold")) +
    xlab("new") + ylab("old") +
    annotate("text", x = -1, y = 2.5, label = (paste0("R^2 = ", lm_cit$adj.r.squared)), size = 5) +
    ggtitle("citrulline_rnt")
print(p1)
```

### Distribution of Rsqaured values from old ~ new lm
```{r distribution of r-squareds, echo=FALSE}
# merge new and old together
dat <- merge(new_raw, old_raw, by = "id_timepoint", all.x = T)
columns_with_metabolite_data <- which(grepl( "compid" , names(dat))==TRUE)
dat <- dat[,c(1, columns_with_metabolite_data)]

# extract the r-squared vals for each metabs
# Define a vector of metabolite names
metabolite_names <- colnames(new_raw[,which(grepl( "compid" , names(new_raw))==TRUE)])

# Create an empty dataframe to store results
rsquared_df <- data.frame(column_name = character(),
                           rsquared = numeric(),
                           stringsAsFactors = FALSE)

# Loop over metabolite names
for (metabolite in metabolite_names) {
  
  # Select data for the metabolite
  dat_metabolite <- na.omit(dat[, grep(paste0(metabolite, "\\.[xy]$"), colnames(dat))])
  
  # Check if there are two columns with the same name
  if (ncol(dat_metabolite) == 2 & nrow(dat_metabolite)>=1) {
    
    # Fit linear model and extract R-squared value
    model <- lm(dat_metabolite[, 2] ~ dat_metabolite[, 1])
    rsquared <- summary(model)$r.squared
    
    # Add results to dataframe
    rsquared_df <- rbind(rsquared_df, data.frame(column_name = metabolite, rsquared = rsquared))
  }
}

# View distribution
hist(rsquared_df$rsquared, breaks = 30, main = "Histogram of R-squareds from lm(old_raw ~ new_raw)")
```


### Distribution of R values from cor(old, new)
```{r distribution of correlations in raw, echo=FALSE}
dat <- merge(new_raw, old_raw, by = "id_timepoint", all.x = T)
columns_with_metabolite_data <- which(grepl( "compid" , names(dat))==TRUE)
dat <- dat[,c(1, columns_with_metabolite_data)]

# Create an empty dataframe to store results
r_df <- data.frame(metabolite = character(),
                   n = numeric(),
                           r = numeric(),
                   p = numeric(),
                           stringsAsFactors = FALSE)

# Define a vector of metabolite names
metabolite_names <- colnames(new_raw[,which(grepl( "compid" , names(new_raw))==TRUE)])

# Loop over metabolite names
for (metabolite in metabolite_names) {
  
  # Select data for the metabolite
  dat_metabolite <- na.omit(dat[, grep(paste0(metabolite, "\\.[xy]$"), colnames(dat))])
  
  # Check if there are two columns with the same name
  if (ncol(dat_metabolite) == 2 & nrow(dat_metabolite)>=3) {
    
    # Calculate correlation coefficient
    r <- cor.test(dat_metabolite[, 1], dat_metabolite[, 2], method = "spearman")
    
    # Add results to dataframe
    r_df <- rbind(r_df, data.frame(metabolite = metabolite, n = nrow(dat_metabolite), r = r$estimate, p = r$p.value))
  }
}

# View distribution
pdf(file = paste0(fig_dir, "Fig4_raw_spearmans_R_histogram.pdf"))
hist(r_df$r, breaks = 30, xlab = "Spearman's correlation coefficient (r)", ylab = "Number of metabolites", main = "")
abline(v = median(r_df$r),                     # Add line for median
       col = "red",
       lwd = 3)
dev.off()
# save out r_df for the raw data - will use this to see if we trust the results for each metab
write.csv(r_df, file = paste0(table_dir, "spearmans_correlation_results.csv"), row.names = F)

# merge in the feature data
r_df_features <- merge(r_df, feature_data, by.x = "metabolite", by.y = "feature_names")
# order by R
r_df_features <- r_df_features[order(r_df_features$r),]
# count the super pathways in those with R <5
table(r_df_features[which(r_df_features$r<0.5),"SUPER_PATHWAY"])
# calculate missingness for each metab
r_df_features$missing <- ((250-r_df_features$n)/250)*100
write.csv(r_df_features, file = paste0(table_dir, "TableS1_pilot_compare.csv"), row.names = F)
```



